---
title: "STAT288 Multiple Regression of FTC Cigarette Data"
author: "Lee Kapp"
date: "5/29/2021"
output: html_document
---

```{r setup, include=FALSE, message=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse); library(skimr); library(gridExtra); library(car); library(corrplot); library(ppcor)
```

### Data Import
```{r import}
cigs<-read.csv("/Users/leekapp/Documents/Portfolio/STAT288CigProject/cigarette.csv")
```

### Descriptive statistics
```{r stats}
sumTar <- summary(cigs$TAR)
sumNic <- summary(cigs$NICOTINE)
sumWt <- summary(cigs$WEIGHT)
sumCO <- summary(cigs$CO)

statsDF <- rbind(Tar = c(round(sumTar[1], 2), round(sumTar[4], 2), round(sd(cigs$TAR), 2), round(sumTar[6], 2)),
                Nicotine = c(round(sumNic[1], 2), round(sumNic[4], 2), round(sd(cigs$NICOTINE), 2), round(sumNic[6], 2)),
                Weight = c(round(sumWt[1], 2), round(sumWt[4], 2), round(sd(cigs$WEIGHT), 2), round(sumWt[6], 2)),
                CO = c(round(sumCO[1], 2), round(sumCO[4], 2), round(sd(cigs$CO), 2), round(sumCO[6], 2)))
statsDF <- as.data.frame(statsDF)
colnames(statsDF) <-  c("Min", "Mean", "Std Dev", "Max")
rownames(statsDF) <- c("Tar (g)", "Nicotine (g)", "Weight (g)", "CO (mg)")
statsDF <- t(statsDF)
statsDF
```


### EDA Plots
```{r EDAplots}
# by weight
p1 <- ggplot(data = cigs, mapping = aes(x = TAR, y = NICOTINE, color = WEIGHT)) +
  geom_point(size = 2)+
  labs(title = "Nicotine vs. Tar Content by Weight", x = "Tar (g)", y = "Nicotine (g)") +
  theme_minimal()

p2 <- ggplot(data = cigs, mapping = aes(x = TAR, y = CO, color = WEIGHT)) +
  geom_point(size = 2)+
  labs(title = "CO vs. Tar Content by Weight", x = "Tar (g)", y = "CO (mg)") +
  theme_minimal()

grid.arrange(p1,p2, nrow=1)

# by tar
p3 <- ggplot(data = cigs, mapping = aes(x = NICOTINE, y = CO, color = TAR)) +
  geom_point(size = 2)+
  labs(title = "CO vs. Nicotine Content by Tar", x = "Nicotine (g)", y = "CO (mg)") +
  theme_minimal()

p4 <- ggplot(data = cigs, mapping = aes(x = WEIGHT, y = CO, color = TAR)) +
  geom_point(size = 2)+
  labs(title = "CO Content vs. Cigarette Weight by Tar", x = "Cigarette weight (g)", y = "CO (mg)") +
  theme_minimal()

p5 <- ggplot(data = cigs, mapping = aes(x = BRAND, y = CO, color = TAR)) +
  geom_point(size = 2)+
  labs(title = "CO yYield by Tar Content", x = "Cigarette Brand", y = "CO (mg)") +
  theme_minimal() +
  theme(axis.text.x = element_text(hjust=1, vjust = 0.25, angle = 90, size = 11))


grid.arrange(p3, p4, nrow=1)
p5

#by nicotine
p6 <- ggplot(data = cigs, mapping = aes(x = TAR, y = CO, color = NICOTINE)) +
  geom_point(size = 2)+
  labs(title = "CO vs. Tar Content by Nicotine", x = "Tar (g)", y = "CO (mg)") +
  theme_minimal()

p7 <- ggplot(data = cigs, mapping = aes(x = BRAND, y = CO, color = NICOTINE)) +
  geom_point(size = 2)+
  labs(title = "CO Yield by Nicotine", x = "Cigarette Brand", y = "CO (mg)") +
  theme_minimal() +
  theme(axis.text.x = element_text(hjust=1, vjust = 0.25, angle = 90, size = 11))

p6
p7

#by CO
p8 <- ggplot(data = cigs, mapping = aes(x = WEIGHT, y = TAR, color = CO)) +
  geom_point(size = 2)+
  labs(title = "Tar Content vs. Weight by CO", x = "Weight (g)", y = "Tar (g)") +
  theme_minimal() #Might have heaviest weight as an outlier

p9 <- ggplot(data = cigs, mapping = aes(x = BRAND, y = TAR, color = CO)) +
  geom_point(size = 2)+
  labs(title = "Tar Content by CO Yield", x = "Cigarette Brand", y = "Tar (g)") +
  theme_minimal() +
  theme(axis.text.x = element_text(hjust=1, vjust = 0.25, angle = 90, size = 11))
p8
p9

```

### Corplot with Bull Durham
```{r corr}
corrplot(cor(cigs[-1]), method = "square", addCoef.col = "red")
# expect multicollinearity between tar and nicotine

# Partial correlation tests with Bull Durham sammple
# Correlation between CO and TAR controlling for Nicotine
pcor.test(cigs$CO, cigs$TAR, cigs["NICOTINE"])
# Correlation between CO and Nicotine controlling for Tar
pcor.test(cigs$CO, cigs$NICOTINE, cigs["TAR"])
# Correlation between CO and Weight controlling for Tar and Nicotine
pcor.test(cigs$CO, cigs$WEIGHT, cigs[,c("TAR","NICOTINE")])
```

### Complete Model
```{r completeMod}
totMod<-(lm(CO~TAR+NICOTINE+WEIGHT, data = cigs))
summary(totMod)
print("Variance Inflation Factors")
vif(totMod)
print("Leverages")
hatvalues(totMod)
summary(hatvalues(totMod))
print("Cook's Distances")
cooks.distance(totMod)

# max values
which.max(hatvalues(totMod))
which.max(cooks.distance(totMod)) #3 is Bull Durham
```

### Tar only model
```{r tarMod}
tarAlone<-(lm(CO~TAR, data = cigs))
summary(tarAlone)
plot(tarAlone) #outputs diagnostic plots - Bull Durham has excessive Cook's distance
ggplot(data = cigs, mapping = aes(y = TAR)) +
  geom_boxplot() +
  labs(title = "Distribution of Tar") +
  theme_minimal()
summary(cigs$TAR)
1.5*(15.1-8.6) + 15.1 #24.85 is max non-outlier for TAR, Bull Durham is 29.8 mg
rstudent(tarAlone) # BullDurham value is -3.722 = an outlier
```

### Nicotine only model
```{r nicMod}
nicAlone<-(lm(CO~NICOTINE, data = cigs))
summary(nicAlone)
```

### Weight only model
```{r wtMod}
wtAlone<-(lm(CO~WEIGHT, data = cigs))
summary(wtAlone)
```

### Nicotine and weight plus interaction term
```{r nicWt}
nicwt<-lm(CO~NICOTINE + WEIGHT + NICOTINE:WEIGHT, data = cigs)
summary(nicwt)

# predictions based on the model with interaction term to show fit in plot
cigs$NWpreds <- predict(nicwt, cigs)

ggplot(data = cigs, mapping = aes(y = CO, x = NICOTINE)) +
  geom_point() +
  geom_smooth(aes(y = NWpreds, x = NICOTINE), method = "lm", se = TRUE, color = "red") +
  labs(title = "CO by Nicotine + Weight + Nicotine:Weight") +
  theme_minimal()

ggplot(data = cigs, mapping = aes(y = hatvalues(nicwt), x = BRAND)) +
  geom_point() +
  theme_minimal() +
  geom_hline(aes(yintercept = mean(hatvalues(nicwt)), linetype = "mean leverage"), color = "red") +
  theme(legend.title = element_blank()) +
  theme(axis.text.x = element_text(hjust=1, vjust = 0.25, angle = 90, size = 11))

hatvalues(nicwt)
meanLevClean<-mean(hatvalues(nicwt))
meanLevClean
par(mfrow = c(2,2))
plot(nicwt)
cooks.distance(nicwt) # Bull Durham = 3.5106
par(mfrow = c(1,1))
plot(rstudent(nicwt)) # Bull Durham = < -3
```

### Tar and weight plus interaction term NOTE: based on correlations, tar and nicotine are basically the same predictors so these models are very similar
```{r tarWt}
tarwt<-lm(CO ~ TAR + WEIGHT + TAR:WEIGHT, data = cigs)
summary(tarwt)
print("Variance Inflation Factors")
vif(tarwt)

# predictions based on the model with interaction term to show fit in plot
cigs$TWpreds <- predict(tarwt, cigs)

ggplot(data = cigs, mapping = aes(y = CO, x = TAR)) +
  geom_point() +
  geom_smooth(aes(y = TWpreds, x = TAR), method = "lm", se = TRUE, color = "red") +
  labs(title = "CO by TAR + Weight + TAR:Weight") +
  theme_minimal()

ggplot(data = cigs, mapping = aes(y = hatvalues(tarwt), x = BRAND)) +
  geom_point() +
  theme_minimal() +
  geom_hline(aes(yintercept = mean(hatvalues(tarwt)), linetype = "mean leverage"), color = "red") +
  theme(legend.title = element_blank()) +
  theme(axis.text.x = element_text(hjust=1, angle = 45, size = 11))

hatvalues(tarwt)
meanLevClean<-mean(hatvalues(tarwt))
meanLevClean
par(mfrow = c(2,2))
plot(tarwt)
cooks.distance(nicwt) # Bull Durham = 3.5106
par(mfrow = c(1,1))
plot(rstudent(nicwt)) # Bull Durham = < -3
```

### Filtering out Bull Durham Sample
```{r noOuts}
cleanCigs<-cigs %>% 
  filter(TAR < 29.8)

cor.test(cleanCigs$CO, cleanCigs$TAR, method = "pearson")
cor.test(cleanCigs$CO, cleanCigs$NICOTINE, method = "pearson")
cor.test(cleanCigs$TAR, cleanCigs$NICOTINE, method = "pearson")
```


### Refitting Without Bull Durham
```{r}
tarWtMod<-lm(CO~TAR + WEIGHT +TAR:WEIGHT, data = cleanCigs)
summary(tarWtMod)

nicWtMod<-lm(CO~NICOTINE + NICOTINE:WEIGHT, data = cleanCigs)
summary(nicWtMod)

taronly<-lm(CO~TAR, data = cleanCigs)
summary(taronly)

tarnic<-lm(CO~TAR + NICOTINE +TAR:NICOTINE, data = cleanCigs)
summary(tarnic)

allthree <- lm(CO~TAR + WEIGHT + NICOTINE, data = cleanCigs)
summary(allthree)
# Nicotine is unnecessary when tar is in the model
print("VIF for the tar + weight + tar:weight model")
vif(tarWtMod)
print("VIF for the nicotine + weight + nicotine:weight model")
vif(nicWtMod)
print("VIF for the tar + nicotine model")
vif(tarnic)
print("VIF for the full model")
vif(allthree)

```

### Corplot Without Bull Durham
```{r corr}
pdf("/Users/leekapp/Documents/Portfolio/STAT288CigProject/corrplot.pdf")
corrplot(cor(cleanCigs[c(2:5)]), method = "square", addCoef.col = "red")
dev.off()
# expect multicollinearity between tar and nicotine

# Partial correlation tests with Bull Durham sammple
# Correlation between CO and TAR controlling for Nicotine
pcor.test(cleanCigs$CO, cleanCigs$TAR, cleanCigs["NICOTINE"])
# Correlation between CO and Nicotine controlling for Tar
pcor.test(cleanCigs$CO, cleanCigs$NICOTINE, cleanCigs["TAR"])
# Correlation between CO and Weight controlling for Tar and Nicotine
pcor.test(cleanCigs$CO, cleanCigs$WEIGHT, cleanCigs[,c("TAR","NICOTINE")])
```

### No interaction terms are significant without the Bull Durham sample
The Multiple R-squared for the tar only model is 0.9335
The Multiple R-squared for the tar + weight + tar:weight model is 0.9359

### Plot of tar only model without Bull Durham sample
```{r simpleMod}
ggplot(data = cleanCigs, mapping = aes(x = TAR, y = CO, color = WEIGHT)) +
  geom_point(size = 2)+
  labs(title = "CO vs. Tar by Weight", x = "Tar (g)", y = "CO (mg)") +
  theme_minimal() +
  geom_smooth(method = 'lm', formula = y~x)
ggsave("/Users/leekapp/Desktop/STAT288CigProject/tarOnly.png")
```

### Tar only model goodness of fit measures
```{r analytics}
ggplot(data = cleanCigs, mapping = aes(x = BRAND, y = taronly$residuals)) +
  geom_point() +
  theme_minimal(base_size = 14) +
  geom_hline(aes(yintercept = 0), color = "red") +
  labs(title = "Residuals for Tar-Only Model", x = "Brand", y = "Residuals") +
  theme(axis.text.x = element_text(hjust = 1, vjust = 0.25, angle = 90, size = 11))
ggsave("/Users/leekapp/Documents/Portfolio/STAT288CigProject/images/tarOnlyResids.png")

ggplot(data = cleanCigs, mapping = aes(x = BRAND, y = hatvalues(taronly))) +
  geom_point() +
  theme_minimal(base_size = 14) +
  geom_hline(aes(yintercept = mean(hatvalues(taronly)), linetype = "mean leverage"), color = "red") +
  labs(title = "Leverages for Tar-Only Model", x = "Brand", y = "Leverages") +
  theme(axis.text.x = element_text(hjust = 1, vjust = 0.25, angle = 90, size = 11)) +
  theme(legend.title = element_blank())
ggsave("/Users/leekapp/Documents/Portfolio/STAT288CigProject/images/tarOnlyLevs.png")

ggplot(data = cleanCigs, mapping = aes(x = BRAND, y = cooks.distance(taronly))) +
  geom_point() +
  theme_minimal(base_size = 14) +
  labs(title = "Cook's Distances for Tar-Only Model", x = "Brand", y = "Cook's Distance") +
  theme(axis.text.x = element_text(hjust = 1, vjust = 0.25, angle = 90, size = 11))
ggsave("/Users/leekapp/Documents/Portfolio/STAT288CigProject/images/tarOnlyCooks.png")
```



